Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-12 main_new.tex
Small changes Simulation study

\begin{itemize}
\item Decisions on where to sample happen sequentially, i.e.  how
  recently acquired information can be assimilated and used to improve
  statistical models and reduce uncertainty associated with where and
  when to make measurements.
\item As a priori information about the state of the ocean is very
  uncertain, there lies a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}, this is
  usually referred to as the \emph{adaptivity gap} \kc{citation
    needed}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the optimal design) and computability (arriving at a
  solution in reasonable time), which has to be considered in
  practice.
\end{itemize}{}

@@

% Determining paths for mobile robotic sensors in order to maximize the
% information gained about an environment is formally known as informative
% path planning. This, and related problems such as the orienteering
% problem \citep{Golden87}, has been studied in the context of graphs,
% where the potential measurement are assigned to nodes on which
% evaluation of different routes can be conducted.
%In this context, a fundamental result from \cite{nemhauser1978analysis} proves that a simple greedy algorithm (iteratively selecting the location which most increases the utility) can achieve a near-optimal solution if the problem can be shown to be \emph{submodular}\footnote{an intuitive diminishing returns property, where the informative value of adding sensors decrease with the number of sensors added.}. Building on this result, \cite{chekuri2005recursive} explored this in the setting of a graph using a recursive greedy algorithm, providing near-optimal solutions depending on the planning horizon and graph resolution. 
Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:sense-plan-act}). Myopic sampling, i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} which focus largely on reducing
predictive variance or mutual information (entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumption of GP models, and the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest by \cite{Low2009} and \cite{fossuminformation}.
In this paper, with the focus on mapping the river plume, we reward
the designs that improve the classification of water masses, as a
means to characterize the frontal zone.

@@

We use a bivariate GP model for temperature and salinity, where we specify the prior mean 
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{a,0} + \beta_{a,1} x_{\mbox{East}} \\
\beta_{a,1} + \beta_{b,1} x_{\mbox{East}}
\end{bmatrix}.
\end{equation}
In this simulation setup, which is motivated by the real case, we expect the eastern part of the domain to be cold freshwater. This situation mimics that of a river mouth entering from the south, and the water masses are pulled to the east. The actual trends are in practiece set from preliminary information or runs of oceanographic models solving the differential equations for fluid flow. In this example the mean is specified by $\beta_{a,1}=0.065$ and $\beta_{b,1}=0.1$. Further, the intercepts of the temperature and salinity variables are set to $\beta_{a,0}=5.8$ and $\beta_{b,1}=29.0$.

@@

\begin{figure}[!h]
  \centering
  \subfigure[Simulated temperature.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_temp.pdf}\label{fig:true_temp}}
  \hfill
  \subfigure[Simulated salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_sal.pdf}\label{fig:true_sal}}
  \hfill
  \subfigure[Excursion set given the realization of temperature and salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_true.pdf}\label{fig:ESet}}
  \hfill
  \subfigure[Excursion probabilities given collected data along the strategy ``static\_north".]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_posterior.pdf}\label{fig:Eprob}}
  \caption{(\ref{fig:true_temp}) and (\ref{fig:true_sal}) shows
    realisations of the simulated temperature and salinity fields, as
    well as the associated excursion set (\ref{fig:ESet}).
    (\ref{fig:Eprob}) show the estimated excursion probabilities after
    performing data collection along the N-S survey line.} 
\label{fig:realisations}
\end{figure}

\subsection{Static and sequential sampling designs}\label{sec:sampling_designs}

@@

We compare three different static designs denoted \texttt{static\_north}, \texttt{static\_east and} \texttt{static\_zigzag} with the three sequential approaches \texttt{naive}, \texttt{myopic} and \texttt{look-ahead}. The static sampling paths are pre-determined, and cannot adapt to data.

For a fixed length, the expected variance after static surveying is analytically available, but here we focus on the properties of the sequential designs which are evaluated using Monte Carlo sampling over several replicates of realizations from the prior, and conducting simulated sequential surveying for each one. For comparison, the same setup is used for the static designs. Figure \ref{fig:Eprob} shows the conditional EP, given data gathered along the north-south survey line for the realization indicated in these displays. In the Monte Carlo replicates, such results are repeatedly computed to approximate the expected variance reduction in the ES. Along with this criterion we also compare predictive performance measured by root mean square error (RMSE) in the temperature and salinity fields and variance reduction in these two field variables. Note that objective function used by the agents is focused on reducing the expected variance in the ES, but we nevertheless hope to achieve good predictive performance for the other criteria as well. Another non-statistical criteria that is important for practical purposes is the computing time of the strategy.

The agent of the sequential designs starts at the center East-West coordinate at the southern end of the domain (node 53), see Fig. \ref{fig:wp_graph}.

@@

At each step, the AUV will choose to move along an edge in this waypoint
graph. While doing so, it gathers three measurements. These are
assimilated in the model, and then a new choice is done at the end of
the edge. The paths of will differ for the various sequential designs.
The designs will also differ between the runs, because the measurements
vary among replicates.

Results of the replicate runs are shown in Figure \ref{fig:sim_results},
where the different criteria are plotted as a function of surveying
distance. Here, expected ES variance in Figure \ref{fig:avg_ev} is
reduced most when using the \texttt{myopic} and \texttt{look-ahead}
strategies. There is not much difference between these two.

>>>>

\begin{itemize}
\item Decisions on where to sample happen sequentially, i.e.  how
  recently acquired information can be assimilated and used to improve
  statistical models and reduce uncertainty associated with where and
  when to make measurements.
\item As a priori information about the state of the ocean is very
  uncertain, there lies a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}, this is
  usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the optimal design) and computability (arriving at a
  solution in reasonable time), which has to be considered in
  practice.
\end{itemize}{}

@@

% Determining paths for mobile robotic sensors in order to maximize the
% information gained about an environment is formally known as informative
% path planning. This, and related problems such as the orienteering
% problem \citep{Golden87}, has been studied in the context of graphs,
% where the potential measurement are assigned to nodes on which
% evaluation of different routes can be conducted.
%In this context, a fundamental result from \cite{nemhauser1978analysis} proves that a simple greedy algorithm (iteratively selecting the location which most increases the utility) can achieve a near-optimal solution if the problem can be shown to be \emph{submodular}\footnote{an intuitive diminishing returns property, where the informative value of adding sensors decrease with the number of sensors added.}. Building on this result, \cite{chekuri2005recursive} explored this in the setting of a graph using a recursive greedy algorithm, providing near-optimal solutions depending on the planning horizon and graph resolution.

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:sense-plan-act}). Myopic sampling, i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} which focus largely on reducing
predictive variance or mutual information (entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumption of GP models, and the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest by \cite{Low2009} and \cite{fossuminformation}.
In this paper, with the focus on mapping the river plume, we reward
the designs that improve the classification of water masses, as a
means to characterize the frontal zone.

@@

We use a bivariate GP model for temperature and salinity, where we specify the prior mean 
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_a(\bx) \\
    \xi_b(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{a,0} + \beta_{a,1} x_{\mbox{East}} \\
\beta_{a,1} + \beta_{b,1} x_{\mbox{East}}
\end{bmatrix}.
\end{equation}
In this simulation setup, which is motivated by the real case, we expect the eastern part of the domain to be cold freshwater. This situation mimics that of a river mouth entering from the south, and the water masses are pulled to the east. The actual trends are in practice set from preliminary information or runs of oceanographic models solving the differential equations for fluid flow. In this example the mean is specified by $\beta_{a,1}=0.065$ and $\beta_{b,1}=0.1$. Further, the intercepts of the temperature and salinity variables are set to $\beta_{a,0}=5.8$ and $\beta_{b,1}=29.0$.

@@

\begin{figure}[!h]
  \centering
  \subfigure[Simulated temperature.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_temp.pdf}\label{fig:true_temp}}
  \hfill
  \subfigure[Simulated salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_sal.pdf}\label{fig:true_sal}}
  \hfill
  \subfigure[Excursion set given the realization of temperature and salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_true.pdf}\label{fig:ESet}}
  \hfill
  \subfigure[Excursion probabilities given collected data along the strategy ``static\_north".]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_posterior.pdf}\label{fig:Eprob}}
  \caption{\ref{fig:true_temp} and \ref{fig:true_sal} shows
    realisations of the simulated temperature and salinity fields, as
    well as the associated excursion set \ref{fig:ESet}.
    \ref{fig:Eprob} show the estimated excursion probabilities after
    performing data collection along the N-S survey line.} 
\label{fig:realisations}
\end{figure}

\newpage
\subsection{Static and sequential sampling designs}\label{sec:sampling_designs}

@@

We compare three different static designs denoted \texttt{static\_north}, \texttt{static\_east and} \texttt{static\_zigzag} with the three sequential approaches \texttt{naive}, \texttt{myopic}, and \texttt{look-ahead}. The static sampling paths are pre-determined and cannot be changed on the basis of data, representing the archetypical pre-planned observation strategies used in current observational practices. For a fixed survey length, the expected variance after static surveying is analytically available (equal coverage each time), but here we focus on the properties from the sequential designs which are evaluated using Monte Carlo sampling over several replicates of realizations from the prior, and conducting simulated sequential surveying for each one; making the result dependent on the observed data. For comparison, the same setup is used for the static designs. Figure \ref{fig:Eprob} shows the conditional EP, given data gathered along the north-south survey line for the realization indicated in these displays. In the Monte Carlo replicates, such results are repeatedly computed to approximate the expected variance reduction in the ES. Along with this criterion we also compare predictive performance measured by root mean square error (RMSE) in the temperature and salinity fields and variance reduction in these two field variables. It is important to note that the objective function used by the agents is focused on reducing the expected variance in the ES, but we nevertheless hope to achieve good predictive performance for criteria such as RMSE as well. Another non-statistical criteria that is important for practical purposes is the computing time of the strategy, as this will impact the performance of the approach in the field for an embedded system.

The sequential sampling agents start at the center East-West coordinate at the southern end of the domain (node 53), see Fig. \ref{fig:wp_graph}.

@@

Maintaining the AUV example, the AUV will move along edges in the Fig. \ref{fig:wp_graph} waypoint graph. While doing so, the AUV will gather measurements. These are assimilated in the GP model, before an evaluation of the next node to sample will be conducted at the end of the edge. The paths of will differ for the various sequential designs due to the individual design criteria and the simulated variability in the environment among the replicates.

Results of the replicate runs are shown in Figure \ref{fig:sim_results},
where the different criteria are plotted as a function of surveying
distance. Here, expected ES variance in Figure \ref{fig:avg_ev} is
reduced most under the \texttt{myopic} and \texttt{look-ahead}
strategies, performing almost equally.
